---
title: "L08 Coords"
subtitle: "Data Visualization (STAT 302)"
author: "Lauren Kennedy"
pagetitle: "L08 Lauren Kennedy"
date: today

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    fig-dpi: 300

execute:
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---


## Overview

The goal of this lab is to use an alternative coordinate system in `ggplot2` to build a plot. 

## Dataset

We'll be using the `mod_nba_adv_stats_2023_24.txt` dataset --- add to your project's `/data` subdirectory. The `mod_nba_adv_stats_2023_24_codebook.txt` provides a quick description of the variables in the dataset --- suggest adding it to the `/data` subdirectory as well. 

```{r}
#| label: load-pkgs-data

# load package(s)
library(tidyverse)
library(janitor)

# load data
nba_dat <- read_delim("data/mod_nba_adv_stats_2023_24.txt", delim = "|") %>%
  janitor::clean_names()

```

## Exercise 

<!-- See the html instructions since they are complete with images and are the official instructions. -->

Using the `mod_nba_adv_stats_2023_24.txt` dataset we will try to recreate/approximate the plot type featured in the [FiveThirtyEight](http://fivethirtyeight.com/) article [Kawhi Leonard Is The Most Well-Rounded Elite Shooter Since Larry Bird](http://fivethirtyeight.com/features/kawhi-leonard-is-the-most-well-rounded-elite-shooter-since-larry-bird/) for any player of your choice for the 2023-2024 season. 

:::{.callout-note icon=false}

## Goal

Replicate the graphic below as closely as possible --- see instructions for image.

:::

### Start with data wrangling

When data wrangling we will need a helper function for creating the quartiles for players. Additionally, it will be useful to exclude players that played less than 10 games or played less than 5 minutes a game. That is, we only want to include "qualified" players.

```{r}
#| label: helper-function

# Returns quartile rank
quartile_rank <- function(x = 0:99) {
  # Set quartile
  quart_breaks <- c(
    -Inf,
    quantile(x,
      probs = c(.25, .5, .75),
      na.rm = TRUE
    ),
    Inf
  )
  cut(x = x, breaks = quart_breaks, labels = FALSE) #cut x into intervals det by quartiles
}
```

:::{.callout-important}
**Add comments to** the code below where indicated. The added comments should concisely describe what the following line(s) of code do in the data wrangling process  
:::

::: {.callout-tip icon=false}

## Solution

```{r}
#| label: data-wrangling

# Graphical data
nba_graph_dat <- nba_dat |> 
  # Filter players who have played in at least 10 games and averaged at least 5 mins/game
  # Keep only 'qualified' players
  filter(g >= 10, mp / g >= 5) |> 
  # Create new columns for the quantile ranks of 5 statistics of interest for each player
  mutate(
    ts_quant = quartile_rank(ts_percent),
    trb_quant = quartile_rank(trb_percent),
    dbpm_quant = quartile_rank(dbpm),
    ast_quant = quartile_rank(ast_percent),
    usg_quant = quartile_rank(usg_percent)
  ) |> 
  # Select player name and the quantile rank columns created above 
  select(player, contains("_quant")) |> 
  # Translate the data in tidy format
  pivot_longer(
    cols = -player, 
    names_to = "variable", 
    values_to = "value"
    ) |> 
  # Sort the data by player name 
  arrange(player)
```

:::

### Build the plot

*Hints:*

- Image width is 8 in, height is 10 in --- use code chunk yaml
- center the image output --- use code chunk yaml
- shrink size down for display to 50% width --- use code chunk yaml
- All added text, except title, is done with `annotate()`
- set `axis.text` to blank in theme
- size 3 and 5 could be useful

::: {.callout-tip icon=false}
## Solution

```{r}
#| label: ex1plot1
#| fig-width: 8
#| fig-height: 10
#| out-width: "50%"
#| fig-align: "center"

player_plot <- "Kelly Oubre Jr."
nba_plot <- nba_graph_dat %>% filter(
  player == player_plot
)


nba_plot %>% ggplot(aes(x = variable, y = value)) +
  geom_col(width = 1, fill = "#F28291") +
  scale_x_discrete(NULL,
                   expand = c(0,0),
                   limits = c("ts_quant", 
                             "usg_quant", 
                             "dbpm_quant",
                             "trb_quant",
                             "ast_quant")) +
  scale_y_continuous(NULL, expand = c(0,0)) +
  geom_hline(yintercept = 1:4, linetype = "dotted") +
  coord_polar(theta = "x") +
  annotate("text",
           label = c("1st-25th", "25th-50th", "50th-75th", "75th-99th"),
           x = rep(3, 4),
           y = c(0.7, 1.7, 2.7, 3.7),
           size = 3) +
  annotate(geom = "text", label = c("TRUE\nSHOOTING",
                                    "USAGE\nRATE",
                                    "DEFENSIVE\nRPM",
                                    "REBOUND\nRATE",
                                    "ASSIST\nRATE"),
          x = seq(1:5), y = rep(5,5),
          size = 5) +
  ggtitle(str_c(player_plot, "(2024)", sep = "\n")) +
  geom_segment(x = 0.5:4.5, y = 0, xend = 0.5:4.5, yend = 4) +
  theme(
    axis.text = element_blank(),
    legend.position = "none"
  )


```


:::

## Challenge

:::{.callout-important icon=false}

Challenge is optional for all students, but we recommend trying it out!

:::

<!-- See the html instructions since they are complete with images and are the official instructions. -->

Continue the work of trying to recreate the graphic from the article. Your final plot should look as follows:

::: {.callout-tip icon=false}
## Solution

:::